Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-11-10 ms/ms/ms.Rmd
<edit ms>

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.

@@

The analysis of morphology by size class demonstrates a significant difference in size between size classes; however, that pattern deviated from our expectations. Elliptical bifaces from Jowell Farm differ in neither shape nor size between the size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Elliptical bifaces in the large and small size classes from Millsap Cache differ significantly in size (RRPP = 10,000, Pr>d = 0.0035), but not shape ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Pairwise comparisons demonstrate that elliptical bifaces from Millsap Cache and Jowell Farm in the large size class are morphologically consistent, but those in the small size class differ in size (RRPP = 10,000, Pr>d = 0.0001) ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Thus, the pattern that emerges is one of differential retouch _intensity_, where isometric reduction strategies of one group (Millsap Cache) may correlate with differing retouch practices that articulate with locally abundant raw material (Kay County flint). This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw material size is insufficient, making it necessary to import the requisite raw material from an extralocal source; in this case, Edwards chert from central Texas.

@@

An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes \citep{RN11107,RN7041,RN5694}. No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache. This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material---resulting in a broader range of biface shapes in the small size class. Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardised (specialised?) approach to reduction practices.

@@

>>>>

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN8605}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. Shape space described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.

@@

The analysis of morphology by size class demonstrates a significant difference in size between size classes; however, that pattern deviated from our expectations. Elliptical bifaces from Jowell Farm differ in neither shape nor size between the size classes ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Elliptical bifaces in the large and small size classes from Millsap Cache differ significantly in size (RRPP = 10,000, Pr>d = 0.0035), but not shape ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)). Pairwise comparisons demonstrate that elliptical bifaces from Millsap Cache and Jowell Farm in the large size class are morphologically consistent, but those in the small size class differ in size (RRPP = 10,000, Pr>d = 0.0001) ([Supplementary Materials](https://seldenlab.github.io/elliptical.bifaces/)).

Thus, the pattern that emerges is one of differential retouch _intensity_, where the isometric retouch strategies of Caddo knappers from Millsap Cache articulate with a different reduction practice thought to be associated with locally abundant raw material (Kay County flint). This stands in contrast to the comparatively minimal retouch intensity of another group (Jowell Farm) where local raw materials were of poor quality and insufficient size, yielding a situation where it was necessary to import raw material from an extralocal source (Edwards chert).

@@

An analysis of morphological disparity was used to identify potential differences in morphological diversity between size classes \citep{RN11107,RN7041,RN5694}. No difference was found for elliptical biface shape or size by site/raw material, but the comparison of morphological diversity by size class demonstrated that biface shapes in the small size class from Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache.

The comparison of morphological disparity by size class demonstrated that elliptical biface shapes in the small size class at Jowell Farm occupy a significantly greater range of morphospace than those from Millsap Cache. This pattern suggests that the comparatively minimal retouch intensity associated with Caddo knappers at Jowell Farm may have been more targeted---potentially due to a lack of abundant local raw material of sufficient quality and size---resulting in a broader range of biface shapes in the small size class. Conversely, these results also suggest that Caddo knappers at Millsap Cache employed a more standardized (specialized?) approach to reduction.

@@

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/elliptical.illustration.png}
\caption{Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region.}
\label{fig:illustration}
\end{figure}

### 2022-11-10 ms/ms/ms.tex
<edit ms>

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Figure
\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

The analysis of morphology by size class demonstrates a significant
difference in size between size classes; however, that pattern deviated
from our expectations. Elliptical bifaces from Jowell Farm differ in
neither shape nor size between the size classes
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Elliptical bifaces in the large and small size classes from
Millsap Cache differ significantly in size (RRPP = 10,000,
Pr\textgreater d = 0.0035), but not shape
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Pairwise comparisons demonstrate that elliptical bifaces
from Millsap Cache and Jowell Farm in the large size class are
morphologically consistent, but those in the small size class differ in
size (RRPP = 10,000, Pr\textgreater d = 0.0001)
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Thus, the pattern that emerges is one of differential
retouch \emph{intensity}, where isometric reduction strategies of one
group (Millsap Cache) may correlate with differing retouch practices
that articulate with locally abundant raw material (Kay County flint).
This stands in contrast to the comparatively minimal retouch intensity
of another group (Jowell Farm) where local raw material size is
insufficient, making it necessary to import the requisite raw material
from an extralocal source; in this case, Edwards chert from central
Texas.

>>>>

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN8605}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Figure
\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

The analysis of morphology by size class demonstrates a significant
difference in size between size classes; however, that pattern deviated
from our expectations. Elliptical bifaces from Jowell Farm differ in
neither shape nor size between the size classes
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Elliptical bifaces in the large and small size classes from
Millsap Cache differ significantly in size (RRPP = 10,000,
Pr\textgreater d = 0.0035), but not shape
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}). Pairwise comparisons demonstrate that elliptical bifaces
from Millsap Cache and Jowell Farm in the large size class are
morphologically consistent, but those in the small size class differ in
size (RRPP = 10,000, Pr\textgreater d = 0.0001)
(\href{https://seldenlab.github.io/elliptical.bifaces/}{Supplementary
Materials}).

Thus, the pattern that emerges is one of differential retouch
\emph{intensity}, where the isometric retouch strategies of Caddo
knappers from Millsap Cache articulate with a different reduction
practice thought to be associated with locally abundant raw material
(Kay County flint). This stands in contrast to the comparatively minimal
retouch intensity of another group (Jowell Farm) where local raw
materials were of poor quality and insufficient size, yielding a
situation where it was necessary to import raw material from an
extralocal source (Edwards chert).
